---
title: "Survival analysis of the ovarian dataset"
output:
  html_document: #default
    toc: true
    #theme: united
    toc_depth: 5
    toc_float: true
    #number_section: true
#editor_options:
  chunk_output_type: console

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 0. Prerequisites: installation and import of libraries
The current work requires the following packages (R version: 3.6.3):

```{r message=FALSE, warning = FALSE}
require(survival)
require(survminer)
require(ranger)
require(ggplot2)
require(magrittr)
require(dplyr)
require(ggfortify)
require(coin)
```


### 1. Exploratory data analysis

Dataset `ovarian` contains information on survival in a randomised trial where two treatments for ovarian cancer were analyzed. Variables:

`futime`:	 survival or censoring time
`fustat`:	 censoring status
`age`:	 in years
`resid.ds`:	 residual disease present (1=no,2=yes)
`rx`:	 treatment group
`ecog.ps`:	 ECOG performance status (1 is better, see reference)

Almsot all variables are supposed to be categorical except for `futime` and `age` which are numerical. 

Take a look at the structure of the data:
```{r message=FALSE, warning = FALSE}
df <- ovarian
str(df)
```


Convert column values into more appropriate data type:
```{r message=FALSE, warning = FALSE}
df <- ovarian
df <- df %>% mutate(ecog.ps = factor(ecog.ps, # ECOG performance status 
                                     levels = c('1', '2'),
                                     labels =  c('better', 'worse')), 
                    
                    resid.ds = factor(resid.ds, # residual disease present
                                      levels = c("1", "2"), 
                                      labels = c("no", "yes")),
                    rx = factor(rx))
str(df)
```


No missing values were found:
```{r message=FALSE, warning = FALSE}
# number of not complete cases
sum(!complete.cases(df))
```


We could also try to transform `age` variable into factor, dividing observations in 2 or more groups. First, take a look at the distribution of age values:
```{r  message=FALSE, warning = FALSE}
ggplot(df, aes(x=age)) + 
  geom_histogram(color="darkblue", fill="lightblue", binwidth=1) +
  
  # add mean value line
  #geom_vline(aes(xintercept=mean(age)),  
   #         color="blue", linetype="dashed", size=1) +
  scale_x_continuous(breaks = seq(round(min(df$age),2), 
                                  round(max(df$age),2), 5)) +
  labs(title = "Distribution of age values in the ovarian data ") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
  
```


Approximately, we could disntiguish 2 groups: 'older'(>=55) and 'younger' (<55) ones:
```{r message=FALSE, warning = FALSE}
df <- df %>% mutate(age_factor = factor(ifelse(age >=55, "older", "younger")))
```



### 2. Kaplan-Meier survival curves

First, let examine if the type of treatment influenced the survival:
```{r message=FALSE, warning = FALSE}
km_trt_fit <- survfit(Surv(futime, fustat) ~ rx, data=df)
autoplot(km_trt_fit) +
  labs(title = "Survival curve depending on the treatment", 
       subtitle = 'Ovarian Cancer Survival Data') +
  xlab('Time') +
  ylab('Survival proportion') +
  theme(axis.text.x = element_text(angle = 90, 
                                     hjust = 1, 
                                     size = 7),
          plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```

We may suggest that the probability to survive is higher in the group with the second tratement than with the first one. 

The effect of age factor in the survival:
```{r message=FALSE, warning = FALSE}
km_age_fit <- survfit(Surv(futime, fustat) ~ age_factor, data=df)
autoplot(km_age_fit) +
  labs(title = "Survival curve depending on age group", 
       subtitle = 'Ovarian Cancer Survival Data') +
  xlab('Time') +
  ylab('Survival proportion') +
  theme(axis.text.x = element_text(angle = 90, 
                                     hjust = 1, 
                                     size = 7),
          plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```


Patients of age less than 55 y.o. have more chances to survive ovarian cancer.

### 3. Log-rank test: Testing for differences in survival between groups


At 5 % signidicance level, there is a difference in survival between age groups:
```{r message=FALSE, warning = FALSE}
survdiff(Surv(futime, fustat) ~ age_factor, data=df)
```

p = 0.03 < 0.05 indicates that comparing survival curves are different in both groups. 

```{r message=FALSE, warning = FALSE}
logrank_test(Surv(futime, fustat) ~ age_factor, data=df)
```


At 5 % signidicance level, there no difference in survival between treatment groups:
```{r message=FALSE, warning = FALSE}
survdiff(Surv(futime, fustat) ~ rx, data=df)
```


```{r message=FALSE, warning = FALSE}
logrank_test(Surv(futime, fustat) ~ rx, data=df)
```


As for presence of residual disease, no significant difference was found (5 % significance level):
```{r message=FALSE, warning = FALSE}
survdiff(Surv(futime, fustat) ~ resid.ds, data=df)
```


```{r message=FALSE, warning = FALSE}
logrank_test(Surv(futime, fustat) ~ resid.ds, data=df)
```

For different ECOG performance status, survival is similar:
```{r message=FALSE, warning = FALSE}
logrank_test(Surv(futime, fustat) ~ ecog.ps, data=df)
```


### 4.  Cox proportional hazards regression analysis


To evaluate simultaneously the effect of several factors on survival, build Cox proportional hazards regression model using the `coxph` function and visualize the result:
```{r message=FALSE, warning = FALSE}
cox <- coxph(Surv(futime, fustat) ~ rx + resid.ds + age_factor + ecog.ps, 
                   data = df)
summary(cox)
```


According to the Wald statistics, treatment group and age group have statistically significant coefficients (at 5 % significance level). Other covariates `ecog.ps` and `resid.ds` failed to have coefficients significantly different from 0. 

Regression coefficients showed that the treatment 2 has lower risk of death than treatment 1 (regression coefficient is `r round(cox$coefficients[[1]], 2)`) and young age group might also have better survival than older age group (coefficient is `r round(cox$coefficients[[3]], 2)`).

Hazard ratios (exponentiated coefficients (exp(coef)) indicated that treatment 2 reduces the hazard by a factor of `r round(exp(unname(cox$coefficients)[1]), 3)` if other covariates are constant. Being of young age reduces the hazard by `r round(exp(unname(cox$coefficients)[3]), 3)` assuming other covariates are fixed. Treatment 2 and young age are associated with good survival. 

At 5 % significance level,  only two overal tests gave significant results: Likelihood ratio test (p=0.02) and Score (logrank) test (p=0.03), but not Wald test (p=0.06). Model might be marginally significant. 

Visualize the estimated survival distribution at the mean values of covariates:
```{r message=FALSE, warning = FALSE}
cox_fit <- survfit(cox)
autoplot(cox_fit)
```


Now let's construct Aalen’s additive regression model for censored data
```{r message=FALSE, warning = FALSE}
aa_fit <- aareg(Surv(futime, fustat) ~ rx + resid.ds + age_factor + ecog.ps, 
                   data = df)

autoplot(aa_fit) +
  labs(title = "Survival curves", 
       subtitle = 'Ovarian Cancer Survival Data') +
  xlab('Time') +
  ylab('Survival proportion') +
  theme(axis.text.x = element_text(angle = 90, 
                                     hjust = 1, 
                                     size = 7),
          plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```

The plots show how the effects of the covariates change over time. Presence of residial disease could increase risk of death over time while younger age and treatment 2 could increase probability of survival. ECOG performance status seems to have the smallest effect on survival. 
We could suggest that age factor is one of the most contributive to survival.   

### 5. Hazard Ratio

```{r message=FALSE, warning = FALSE}
ggforest(cox, data = df)
```

Being of younger age and treated with the treatment 2 could  signficiantly reduce the risk of death/increase the probability of survival: at any given time, 0.17 times as many young patients could die as older ones and 0.28 times as many treated with the treatment 2 could pass away as those treated with the treatment 1.
Contribution of other factors to survival are not significant at 5 % significance level. 

